prolfqua&colon; A Comprehensive R-package for Proteomics Differential Expression Analysis

### 2022-10-31 inst/MyArticle/paper.Rmd
Changes related to the manuscript resubmission ACS JPR.

---
title: prolfqua&colon; A Comprehensive R-package for Proteomics Differential Expression Analysis
author: 
  - Witold E. Wolski $^{\dagger\ddagger}$^[[Functional Genomics Center Zurich (FGCZ)](https://fgcz.ch) - [University of Zurich](https://uzh.ch)/[ETH Zurich](https://ethz.ch), Winterthurerstrasse 190, CH-8057 Zurich, Switzerland. $^\ddagger$[Swiss Institute of Bioinformatics (SIB), Quartier Sorge - Batiment Amphipole, 1015 Lausanne, Switzerland](https://www.sib.swiss/) $^\dagger$Correspondence&colon; \texttt{wew@fgcz.ethz.ch}]
  - Paolo Nanni$^{\ast}$
  - Jonas Grossmann$^{\ast\ddagger}$
  - Maria d'Errico$^{\ast\ddagger}$
  - Ralph Schlapbach$^{\ast}$
  - Christian Panse$^{\ast\ddagger}$
output: 
  bookdown::pdf_document2:
      toc: true
      toc_appendix: false
      toc_bib: false
      keep_tex: true
      citation_package: natbib
date: "`r format(Sys.time())`"
bibliography: prolfqua.bib
vignette: >
  %\VignetteIndexEntry{prolfqua&colon; A Comprehensive R-package for Proteomics Differential Expression Analysis} 
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
abstract: |
    Mass spectrometry is widely used for quantitative proteomics studies, relative protein quantification, and differential expression analysis of proteins. There is a large variety of quantification software and analysis tools. Nevertheless, there is a need for a modular, easy-to-use application programming interface in R that transparently supports a variety of well principled statistical procedures to make applying them to proteomics data, comparing and understanding their differences easily. The `r BiocStyle::Githubpkg("fgcz/prolfqua")` package integrates essential steps of the mass spectrometry-based differential expression analysis workflow: quality control, data normalization, protein aggregation, statistical modeling, hypothesis testing, and sample size estimation. The package makes integrating new data formats easy. It can be used to model simple experimental designs with a single explanatory variable and complex experiments with multiple factors and hypothesis testing. The implemented methods allow sensitive and specific differential expression analysis. Furthermore, the package implements benchmark functionality that can help to compare data acquisition, data preprocessing, or data modeling methods using a gold standard dataset. The application programmer interface of `r BiocStyle::Githubpkg("fgcz/prolfqua")`  strives to be clear,  predictable, discoverable, and consistent to make proteomics data analysis application development easy and exciting. Finally, the `r BiocStyle::Githubpkg("fgcz/prolfqua")` R package is available on GitHub https://github.com/fgcz/prolfqua, distributed under the MIT license. It runs on all platforms supported by the [R free software environment for statistical computing and graphics](https://www.r-project.org/).

@@

\begin{quote}
To paraphrase provocatively, `machine learning is statistics minus any checking of models and assumptions'.
\begin{flushright}
-- Brian D. Ripley, useR! 2004, Vienna.
\end{flushright}
\end{quote}

@@

Several data analysis packages exist to model mass spectrometry protein quantification experiments, e.g., `r Biocpkg("limma")` [@Ritchie2015], `r Biocpkg("MSstats")` [@Choi2014], `r Biocpkg("PECA")` [@Suomi2017bEnhanced],  `r Biocpkg("msqrob2")` [@Goeminne2016] or `r Biocpkg("proDA")` [@bioxrvproDA2020], to mention some, all implemented in R. Each of them has some unique features, for example, *MSstats* determines the statistical model from the structure of the sample annotation, which allows users with limited statistical knowledge to perform differential expression analysis (DEA). At the same time, *limma* allows the specification of a design matrix using a linear model formula and implements the empirical Bayes variance shrinkage method. The package *PECA* performs a roll-up of peptide level differences and peptide level $p$-value estimates obtained from *limma* or *PECA*, to protein level estimates. Furthermore, *msqrob2* combines two models: robust linear models fitted to protein abundances and a quasibinomial generalized linear model fitted to peptide counts, into Hurdle model. The *proDA* package implements a linear probabilistic dropout model to account for missing data without imputation.

Of note are the various approaches to handle missing observations, which are inherent in mass spectrometric bottom-up experiments. For instance, *MSstats* handles missing data by feature filtering and imputation. Other means of modeling missing observations are the Hurdle models discussed by @goeminne2020msqrob, while the R package *proDA* models missingness using probabilistic dropout models [@bioxrvproDA2020].

We can use all the R packages discussed when analyzing parallel-group designs using a single explanatory variable and contrasting groups. However, we can use only some of them if we want to model factorial designs or repeated measurements. Table \@ref(tab:tableOverview) gives an overview of the models and features supported by these packages. We see that, for instance, *limma* and *proDA* allow us to fit a comprehensive variety of models and test various hypotheses; however, good knowledge of the design matrix specification using the R formula interface is required [@law2020guide].

@@

When developing the R package `r BiocStyle::Githubpkg("fgcz/prolfqua")` we were inspired by the R package `r BiocStyle::CRANpkg("caret")` [@caret2008] which enables us to call various machine learning (ML) methods, and makes selecting the best ML algorithm for your problem easy. We aimed for a similar R package for the DEA of quantitative proteomics data. However, the existing packages differ widely regarding supported designs, model specifications, and output formats. At the same time, they have the following features in common: fitting linear models to either peptide or protein intensities, determining differences among groups, and afterward applying empirical Bayes variance shrinkage. Therefore, the revised goal was to provide a modular object-oriented design, with R linear models as a core, and add features such as $p$-value aggregation, variance shrinkage, or modeling of missing observations.

Furthermore, `r BiocStyle::Githubpkg("fgcz/prolfqua")` also includes methods specific to proteomics data. For example, we implemented strategies to estimate protein intensities from peptide intensities: top N [@Grossmann2010], Tukey's median polish  [@tukey1977exploratory], robust linear models [@goeminne2020msqrob]. Peptide or protein abundances can then be scaled and transformed using robust scaling, _quantile_ normalization or `r Biocpkg("vsn")` to remove systematic differences among samples and heteroscedasticity. In this respect, *prolfqua* can be compared with R packages such as `r Biocpkg("DEP")` [@DEP2018] or `r Biocpkg("POMA")` [@POMA2021] which support the entire 
DEA pipeline.

Since group sizes are relatively small, typically with four or five subjects per group, the high power of the tests is a relevant criterion to assess the modeling method. The quantified proteins can be ranked using the estimated fold-change, $t$-statistics, or scaled $p$-value, and afterward subjected to gene set enrichment (GSEA) or over-representation analysis [@subramanian2005gene] to determine up or down-regulated groups of proteins. Furthermore, the statistical model must provide an unbiased estimate of the false discovery rate (FDR) to manage expectations when selecting protein lists for follow-up experiments. We will use the partial area under the receiver operator curve (ROC) to assess the power of the tests and compare the FDR with the false discovery proportion ($FDP$). We use the *IonStar*[@shen2018ionstar] and *CPTAC*[@edwards2015cptac]  datasets, processed with *MaxQuant* and *FragPipe* to benchmark the modeling methods implemented in *prolfqua* and to compare our results with those of *MSstats*, *msqrob2* and *proDA*.

Although other benchmark datasets exist[@wessels2012comprehensive;@o2018proteome], the *IonStar* dataset has the advantage that the expected differences among groups of the spike in proteins are slight compared to other benchmark datasets, making DEA more difficult and enabling us to see performance differences. For the same reason, of making the DEA more challenging and enhancing the differences between the statistical modeling approaches, we focus on the *MaxQuant* outputs since they are noisier than those produced by *FragPipe*.

@@

We store all the data needed for analysis in a data frame as tidy data, i.e., every column is a variable, every row is an observation, and every cell is a single value [@tidydata2014]. Using an R6 [@R6cite] configuration object (Figure \@ref(fig:LFQData)), we specify what variable is in which column making it easy to integrate new inputs in *prolfqua* if provided as a tidy data. For example, to visualize tidy Spectronaut[@bruderer2015extending], DiaNN[@demichev2020dia], or Skyline[@maclean2010skyline] outputs, or data in *MSstats*[@Choi2014] format, only a few lines of code are needed to update the *prolfqua* `AnalysisTableConfiguration` configuration. The configuration encapsulates the differences among the various input formats in column names and enables the use of the methods without additional arguments. An example code for creating an *FragPipe*[@yu2020fast] configuration can be found in the Supplementary Section: "Creating a prolfqua configuration". We implemented methods that transform the data into tidy data for popular software like *MaxQuant*[@Cox2008MaxQuant], or *FragPipe*, which stores the same variable, e.g., intensity, in multiple columns, one for each sample. Relying on the tidy data table enables us to interface with many data manipulation, visualization, and modeling methods, implemented in base R [@RCoreTeam2021] and the tidyverse [@tidyverse2019], easily. We use R6 classes to structure the functionality of the package (see Figure \@ref(fig:LFQData) and Figure \@ref(fig:ContrastUML)). R6 classes are well supported by command-line completion features (see Supplementary Figure 8) in *RStudio* [@RStudio2022], and help to implement argument free functions.

(ref:LFQData) Class Diagram of classes representing the proteomics data. The `LFQData` class encapsulates the quantitative proteomics data stored in a table of tidy data. An instance of the AnalysisTableConfiguration class specifies a mapping of table columns to sample names, peptide or protein identifiers, explanatory variables, and response variables. The `LFQDataPlotter` class and other classes decorate the `LFQData` class with additional functionality. Similarly, the `LFQDataStats` and `LFQDataSummary` reference the `LFQData` class and group methods for variance and sample size estimation or summarizing peptide and protein counts.

@@

R's formula interface for linear models is flexible, widely used, and well documented [@faraway2016extending; @law2020guide]. We use the formula interface to specify the models, making it easy to reproduce an analysis performed with *prolfqua* in other statistical programming languages. In addition, we implement features specific to high throughput experiments, such as the empirical Bayes variance and $p$-value moderation, which utilizes the parallel structure of the protein measurements and the analysis [@Ritchie2015]. We also compute probabilities of differential protein regulation based on peptide level models [@Suomi2017bEnhanced]. We used R6 classes to encapsulate the statistical modelling functionality in *prolfqua* (see Figure \@ref(fig:ContrastUML)). We specify a contrast interface (`ContrastsInterface`). Several implementations allow to perform DEA given linear or mixed effect models (`Contrasts`), variance shrinkage (`ContrastsModerated`), or to impute contrasts in cases when observations are missing for an entire group of samples (`ContrastsSimpleImpute`). Further implementations of the interface encapsulate and integrate DEA results of external tools such as *proDA* or of *SAINTexpress* [@teo2014saintexpress] used to analyze data from protein interaction studies.

@@

To evaluate the performance of DEA, we use the *IonStar* benchmark dataset[@shen2018ionstar], available from the Proteomics Identifications Database (PRIDE) identifier PXD003881. $DH5\alpha$ _E. coli_ lysate was spiked in human pancreatic cancer cells (Panc-1) lysate at five different levels: $3\%$, $4.5\%$, $6\%$, $7.5\%$ and $9\%$ _E. coli_. We annotated these dilutions from smallest to largest with the letters _a_, _b_, _c_, _d_ and _e_. By comparing the various dilutions, we obtain different effect sizes, e.g., when comparing dilution _e_ (9%) against dilution _d_ (7.5%), the expected difference is $1.2$ for E. coli proteins and $1$ for human proteins. There are four technical replicates for each dilution, hence 20 raw files in total. To compare the performance of the various methods implemented in *prolfqua* we use only the contrasts resulting in minor differences $\Delta = (1.20, 1.25, 1.30, 1.50)$, because for bigger differences all models perform similarly.

@@

We processed the raw data of the *IonStar* dataset using *MaxQuant* [@Cox2008MaxQuant] version Version 1.6.10.43, with *MaxQuant* default settings for orbitrap data. The text files generated by *MaxQuant* are available in the *prolfquadata* R package [@prolfquadata]. *MaxQuant* produces various output files which can be used for DEA. We are using the quantification results reported in the 'peptide.txt' file for DEA. However, *MSstats* is using the 'evidence.txt' file for the DEA.

@@

We processed the raw data of the *IonStar* dataset using the *FragPipe* [@yu2020fast] Version 14, with the default workflow for label free quantification with match between runs. The text files generated by *FragPipe* are available in the *prolfquadata* R package [@prolfquadata].
Similarily to *MaxQuant*, the *FragPipe* software produces various outputs which can be used for DEA. We used the total protein intensities reported in the 'combined_protein.tsv' file as input for the DEA and call this dataset IonStar/FragPipe/combined_protein.tsv. Alternatively, we benchmarked the DEA using the 'MSstats.tsv' file as input and call this dataset IonStar/FragPipe/combined_protein.tsv.

@@

We used the CPTAC dataset, available in the R package *msdata*, and described in [@Goeminne2016]. In brief, the Sigma Universal Protein Standard mixture 1 (UPS1) containing $48$ different human proteins was spiked in a protein background of 60 ng/$\mu L$ Saccharomyces cerevisiae strain BY4741. Two different spike-in concentrations were used, 6A ($0.25$ fmol UPS1 proteins/$\mu L$) and 6B ($0.74$ fmol UPS1 proteins/$\mu L$). Three replicates are available for each concentrations. The data were searched with *MaxQuant* version `1.5.2.8.`.

@@

with $X$ being the design matrix. The degrees of freedom for the contrast are equal to the residual degrees of freedom of the linear model. For estimating contrasts from mixed effects models we used the function `contest` implemented in the R package `r BiocStyle::CRANpkg("lmerTest")`  and used the Satterthwaite [@Kuznetsova2017lmerTest] method to estimate the denominator degrees of freedom. These methods are available in the class `Contrast` (see Figure \@ref(fig:ContrastUML))

@@

# Results and Discussion.

@@

The code snippets in this section demonstrate how a DEA workflow can be implemented using the *prolfqua* R package. To speed up the computation of these examples, we use a subset of the Ionstar dataset generated by randomly selecting $400$ proteins. First, we remove all proteins with a single peptide and all observations for which *MaxQuant* reports zero intensities which leaves $332$ proteins. Next, peptide abundances are $\log_2$ transformed and robust z-score scaled using the method `robscale`. Then, using the `LFQDataPlotter` class, we show the distribution of the normalized peptide abundances in Figure \@ref(fig:figure1) Panel A. Afterwards, protein intensities are estimated from peptide intensities using Tukey's median polish. Figure \@ref(fig:figure1) Panel B shows the peptide intensities and the estimated protein intensities. Next, we compute the standard deviation of all the proteins in each group and display their distribution using violin plots (Panel C). Finally, we create a boxplot (Panel D) showing the abundance of one protein.

@@

(ref:figure2) Panel A - Histogram showing the distribution of p-values for $332$ proteins. Panel B -  Volcano plot showing $-\log_{10}$ transformed $FDR$ as a function of the difference between groups for $332$ proteins. With black dots, we show effect size and FDR estimates obtained from the linear model, while in green, we plot those obtained using imputation. Finally, panel C shows the difference between groups as a function of the rank of the abundance of the proteins.

```{r figure2, fig.cap="(ref:figure2)", echo=TRUE, out.width = '90%', fig.width = 8, fig.height = 5}
## specify differences among groups
contrasts <- c(
  "dilution_(9/7.5)_1.2" =   "dilution.e - dilution.d",
  "dilution_(7.5/6)_1.25" =   "dilution.d - dilution.c"
)
## fit model
lmmodel <- paste(lfqpro$response()," ~ dilution.")
modelFunction <- prolfqua::strategy_lm( lmmodel, model_name = "lm")
models <- prolfqua::build_model(lfqpro, modelFunction)

@@

R linear model and linear mixed effect models allow modeling parallel designs, repeated measurements, factorial designs, and many more. Models in *prolfqua* are specified using R's linear model and mixed model formula interface. Therefore, knowledge of the R regression model infrastructure [@faraway2016extending; @MASS2002] is advantageous when using our package. Furthermore, this glass box approach should make it easy to reimplement an analysis performed with *prolfqua* using base R or other programming languages by reading the analysis script.
However, in the package documentation, we showcase how a user, without this knowledge, can analyze experiments with a parallel group design and a factorial design.

@@

To ease the usage barriers of the R package to users not familiar with statistics and R programming, we developed an application based on the *prolfqua* package into our data management platform [B-Fabric](https://fgcz-bfabric.uzh.ch/) [@bfabric;@bfabricjib2022]. The B-Fabric system runs a computing infrastructure controlled by a local resource management system that supports cloud-bursting [@VMMAD]. This integration enables users to select the input data and basic settings in a graphical user interface (GUI).

In this GUI application, we determine the model formula, and which comparisons to compute from the sample annotation provided in tabular form, similarly to *SAINTexpress* or *MSstats*. However, this application can analyze only parallel group designs with and without repeated measurements or factorial designs without interactions. The user receives a report including input files, the R markdown file, and R scripts necessary to replicate the analysis using their in-house R installation. In this way, *prolfqua* and [B-Fabric](https://fgcz-bfabric.uzh.ch/) and help scientists to meet requirements from funding agencies, journals, and academic institutions while publishing their data according to the FAIR  [@FAIR] data principles. We are working on creating a shiny stand-alone application with the described functionality and making it available soon.

@@

The Benchmark functionality of *prolfqua* includes receiver operator curves (ROC) and computes partial areas under those curves ($pAUC$) and other scores, e.g., the false discovery proportion $FDP$. We use those scores, i.e. the $pAUC$ at $10%$ $FDR$ and the $FDP$, to examine how well the methods implemented in *prolfqua* model quantitative mass spectrometric high throughput data and compare them with the results produced by *MSstats*, *proDA* and *msqrob2*. Table \@ref(tab:describeBenchmarkedModels) summarizes all the modelling methods we evaluated.

@@

Using a benchmark dataset whose ground truth is known (see Methods), we assessed the performance of different modeling approaches implemented in *prolfqua*, *MSstats*, *proDA* and *msqrob2*. Table \@ref(tab:describeBenchmarkedModels) summarizes which methods we have evaluated, which *MaxQuant* files we used as input, and if the models are fitted to peptide or protein intensities. We make the R-markdown files to replicate the benchmarking available at `r BiocStyle::Githubpkg("wolski/prolfquabenchmark")`.

>>>>

---
title: prolfqua&colon; A Comprehensive R-package for Proteomics Differential Expression Analysis
author: 
  - Witold E. Wolski $^{\dagger\ddagger}$^[[Functional Genomics Center Zurich (FGCZ)](https://fgcz.ch) - [University of Zurich](https://uzh.ch)/[ETH Zurich](https://ethz.ch), Winterthurerstrasse 190, CH-8057 Zurich, Switzerland. $^\ddagger$[Swiss Institute of Bioinformatics (SIB), Quartier Sorge - Batiment Amphipole, 1015 Lausanne, Switzerland](https://www.sib.swiss/) $^\dagger$Correspondence&colon; \texttt{wew@fgcz.ethz.ch}]
  - Paolo Nanni$^{\ast}$
  - Jonas Grossmann$^{\ast\ddagger}$
  - Maria d'Errico$^{\ast\ddagger}$
  - Ralph Schlapbach$^{\ast}$
  - Christian Panse$^{\ast\ddagger}$
output: 
  bookdown::pdf_document2:
      toc: true
      toc_appendix: false
      toc_bib: false
      keep_tex: true
      citation_package: natbib
date: "`r format(Sys.time())`"
bibliography: prolfqua.bib
vignette: >
  %\VignetteIndexEntry{prolfqua&colon; A Comprehensive R-package for Proteomics Differential Expression Analysis} 
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
abstract: |
    Mass spectrometry is widely used for quantitative proteomics studies, relative protein quantification, and differential expression analysis of proteins. There is a large variety of quantification software and analysis tools. Nevertheless, there is a need for a modular, easy-to-use application programming interface in *R* that transparently supports a variety of well principled statistical procedures to make applying them to proteomics data, comparing and understanding their differences easily. The `r BiocStyle::Githubpkg("fgcz/prolfqua")` package integrates essential steps of the mass spectrometry-based differential expression analysis workflow: quality control, data normalization, protein aggregation, statistical modeling, hypothesis testing, and sample size estimation. The package makes integrating new data formats easy. It can be used to model simple experimental designs with a single explanatory variable and complex experiments with multiple factors and hypothesis testing. The implemented methods allow sensitive and specific differential expression analysis. Furthermore, the package implements benchmark functionality that can help to compare data acquisition, data preprocessing, or data modeling methods using a gold standard dataset. The application programmer interface of `r BiocStyle::Githubpkg("fgcz/prolfqua")`  strives to be clear,  predictable, discoverable, and consistent to make proteomics data analysis application development easy and exciting. Finally, the `r BiocStyle::Githubpkg("fgcz/prolfqua")` *R* package is available on GitHub https://github.com/fgcz/prolfqua, distributed under the MIT license. It runs on all platforms supported by the [*R* free software environment for statistical computing and graphics](https://www.r-project.org/).

@@

@@

Several data analysis packages exist to model mass spectrometry protein quantification experiments, e.g., `r Biocpkg("limma")` [@Ritchie2015], `r Biocpkg("MSstats")` [@Choi2014], `r Biocpkg("PECA")` [@Suomi2017bEnhanced],  `r Biocpkg("msqrob2")` [@Goeminne2016] or `r Biocpkg("proDA")` [@bioxrvproDA2020], to mention some, all implemented in *R* [@RCoreTeam2021]. Each of them has some unique features, for example, *MSstats* determines the statistical model from the structure of the sample annotation, which allows users with limited statistical knowledge to perform differential expression analysis (DEA). At the same time, *limma* allows the specification of a design matrix using a linear model formula and implements the empirical Bayes variance shrinkage method. The package *PECA* performs a roll-up of peptide level differences and peptide level $p$-value estimates obtained from *limma* or *PECA*, to protein level estimates. Furthermore, *msqrob2* combines two models: robust linear models fitted to protein abundances and a quasibinomial generalized linear model fitted to peptide counts, into Hurdle model. The *proDA* package implements a linear probabilistic dropout model to account for missing data without imputation.

Of note are the various approaches to handle missing observations, which are inherent in mass spectrometric bottom-up experiments. For instance, *MSstats* handles missing data by feature filtering and imputation. Other means of modeling missing observations are the Hurdle models discussed by @goeminne2020msqrob, while the *R* package *proDA* models missingness using probabilistic dropout models [@bioxrvproDA2020].

We can use all the *R* packages discussed when analyzing parallel-group designs using a single explanatory variable and contrasting groups. However, we can use only some of them if we want to model factorial designs or repeated measurements. Table \@ref(tab:tableOverview) gives an overview of the models and features supported by these packages. We see that, for instance, *limma* and *proDA* allow us to fit a comprehensive variety of models and test various hypotheses; however, good knowledge of the design matrix specification using the *R* formula interface is required [@law2020guide].

@@

When developing the *R* package `r BiocStyle::Githubpkg("fgcz/prolfqua")` we were inspired by the *R* package `r BiocStyle::CRANpkg("caret")` [@caret2008] which enables us to call various machine learning (ML) methods, and makes selecting the best ML algorithm for your problem easy. We aimed for a similar *R* package for the DEA of quantitative proteomics data. However, the existing packages differ widely regarding supported designs, model specifications, and output formats. At the same time, they have the following features in common: fitting linear models to either peptide or protein intensities, determining differences among groups, and afterward applying empirical Bayes variance shrinkage. Therefore, the revised goal was to provide a modular object-oriented design, with *R* linear models as a core, and add features such as $p$-value aggregation, variance shrinkage, or modeling of missing observations.

Furthermore, `r BiocStyle::Githubpkg("fgcz/prolfqua")` also includes methods specific to proteomics data. For example, we implemented strategies to estimate protein intensities from peptide intensities: top N [@Grossmann2010], Tukey's median polish  [@tukey1977exploratory], robust linear models [@goeminne2020msqrob]. Peptide or protein abundances can then be scaled and transformed using robust scaling, _quantile_ normalization or `r Biocpkg("vsn")` to remove systematic differences among samples and heteroscedasticity. In this respect, *prolfqua* can be compared with *R* packages such as `r Biocpkg("DEP")` [@DEP2018] or `r Biocpkg("POMA")` [@POMA2021] which support the entire 
DEA pipeline.

Since group sizes are relatively small, typically with four or five subjects per group, the high power of the tests is a relevant criterion to assess the modeling method. The quantified proteins can be ranked using the estimated fold-change, $t$-statistics, or scaled $p$-value, and afterward subjected to gene set enrichment (GSEA) or over-representation analysis [@subramanian2005gene] to determine up or down-regulated groups of proteins. Furthermore, the statistical model must provide an unbiased estimate of the false discovery rate (FDR) to manage expectations when selecting protein lists for follow-up experiments. We will use the partial area under the receiver operator curve (ROC) to assess the power of the tests and compare the FDR with the false discovery proportion ($FDP$). We use the *IonStar*[@shen2018ionstar] and *CPTAC*[@edwards2015cptac]  datasets, processed with *MaxQuant* and *FragPipe* to benchmark the modeling methods implemented in *prolfqua* and to compare our results with those of *MSstats*, *msqrob2* and *proDA*. Although other benchmark datasets exist[@wessels2012comprehensive;@o2018proteome], the *IonStar* dataset has the advantage that the expected differences, for the spike in proteins, among groups are small compared to other benchmark datasets, making DEA more difficult and enabling us to see performance differences among the modeling methods.

@@

We store all the data needed for analysis in a data frame as tidy data, i.e., every column is a variable, every row is an observation, and every cell is a single value [@tidydata2014]. Using an R6 [@R6cite] configuration object (Figure \@ref(fig:LFQData)), we specify what variable is in which column making it easy to integrate new inputs in *prolfqua* if provided as a tidy data. For example, to visualize tidy Spectronaut[@bruderer2015extending], DiaNN[@demichev2020dia], or Skyline[@maclean2010skyline] outputs, or data in *MSstats*[@Choi2014] format, only a few lines of code are needed to update the *prolfqua* `AnalysisTableConfiguration` configuration. The configuration encapsulates the differences among the various input formats in column names and enables the use of the methods without additional arguments. An example code for creating an *FragPipe*[@yu2020fast] configuration can be found in the Supplementary Section "Creating a prolfqua configuration". We implemented methods that transform the data into tidy data for popular software like *MaxQuant*[@Cox2008MaxQuant], or *FragPipe*, which stores the same variable, e.g., intensity, in multiple columns, one for each sample. Relying on the tidy data table enables us to interface with many data manipulation, visualization, and modeling methods, implemented in base *R* [@RCoreTeam2021] and the tidyverse [@tidyverse2019], easily. We use R6 classes to structure the functionality of the package (see Figure \@ref(fig:LFQData) and Figure \@ref(fig:ContrastUML)). R6 classes are well supported by command-line completion features (see Supplementary Figure 8) in *RStudio* [@RStudio2022], and help to implement argument free functions.

(ref:LFQData) Class Diagram of classes representing the proteomics data. The `LFQData` class encapsulates the quantitative proteomics data stored in a table of tidy data. An instance of the AnalysisTableConfiguration class specifies a mapping of table columns to sample names, peptide or protein identifiers, explanatory variables, and response variables. The `LFQDataPlotter` class and other classes decorate the `LFQData` class with additional functionality. For instance, the `LFQDataStats` and `LFQDataSummary` reference the `LFQData` class and group methods for variance and sample size estimation or summarizing peptide and protein counts. Furthermore, the `LFQDataTransformer` and `LFQDataAggregator` classes group functions for data normalization and for estimating protein from peptide intensities.

@@

*R*'s formula interface for linear models is flexible, widely used, and well documented [@faraway2016extending; @law2020guide]. We use the formula interface to specify the models, making it easy to reproduce an analysis performed with *prolfqua* in other statistical programming languages. In addition, we implement features specific to high throughput experiments, such as the empirical Bayes variance and $p$-value moderation, which utilizes the parallel structure of the protein measurements and the analysis [@Ritchie2015]. We also compute probabilities of differential protein regulation based on peptide level models [@Suomi2017bEnhanced]. We used R6 classes to encapsulate the statistical modelling functionality in *prolfqua* (see Figure \@ref(fig:ContrastUML)). We specify a contrast interface (`ContrastsInterface`). Several implementations allow to perform DEA given linear or mixed effect models (`Contrasts`), variance shrinkage (`ContrastsModerated`), or to impute contrasts in cases when observations are missing for an entire group of samples (`ContrastsSimpleImpute`). Further implementations of the interface encapsulate and integrate DEA results of external tools such as *proDA* or of *SAINTexpress* [@teo2014saintexpress] used to analyze data from protein interaction studies.

@@

To evaluate the performance of DEA, we use the *IonStar* benchmark dataset[@shen2018ionstar], available from the Proteomics Identifications Database (PRIDE) identifier PXD003881. $DH5\alpha$ _E. coli_ lysate was spiked in human pancreatic cancer cells (Panc-1) lysate at five different levels: $3\%$, $4.5\%$, $6\%$, $7.5\%$ and $9\%$ _E. coli_. We annotated these dilutions from smallest to largest with the letters _a_, _b_, _c_, _d_ and _e_. By comparing the various dilutions, we obtain different effect sizes, e.g., when comparing dilution _e_ ($9\%$) against dilution _d_ ($7.5\%$), the expected difference is $1.2$ for E. coli proteins and $1$ for human proteins. There are four technical replicates for each dilution, hence 20 raw files in total. To compare the performance of the various methods implemented in *prolfqua* we use only the contrasts resulting in minor differences $\Delta = (1.20, 1.25, 1.30, 1.50)$, because for bigger differences all models perform similarly.

@@

We processed the raw data of the *IonStar* dataset using *MaxQuant* [@Cox2008MaxQuant] version Version 1.6.10.43, with *MaxQuant* default settings for orbitrap data. The text files generated by *MaxQuant* are available in the *prolfquadata* *R* package [@prolfquadata]. *MaxQuant* produces various output files which can be used for DEA. We are using the quantification results reported in the 'peptide.txt' file for DEA. However, *MSstats* is using the 'evidence.txt' file for the DEA.

@@

We processed the raw data of the *IonStar* dataset using the *FragPipe* [@yu2020fast] Version 14, with the default workflow for label free quantification with match between runs. The text files generated by *FragPipe* are available in the *prolfquadata* *R* package [@prolfquadata].
Similarily to *MaxQuant*, the *FragPipe* software produces various outputs which can be used for DEA. We used the total protein intensities reported in the 'combined_protein.tsv' file as input for the DEA and call this dataset IonStar/FragPipe/combined_protein.tsv. Alternatively, we benchmarked the DEA using the 'MSstats.tsv' file as input and call this dataset IonStar/FragPipe/combined_protein.tsv.

@@

We used the CPTAC dataset, available in the *R* package *msdata*, and described in [@Goeminne2016]. In brief, the Sigma Universal Protein Standard mixture 1 (UPS1) containing $48$ different human proteins was spiked in a protein background of 60 ng/$\mu L$ Saccharomyces cerevisiae strain BY4741. Two different spike-in concentrations were used, 6A ($0.25$ fmol UPS1 proteins/$\mu L$) and 6B ($0.74$ fmol UPS1 proteins/$\mu L$). Three replicates are available for each concentrations. The data were searched with *MaxQuant* version `1.5.2.8.`.

@@

with $X$ being the design matrix. The degrees of freedom for the contrast are equal to the residual degrees of freedom of the linear model. For estimating contrasts from mixed effects models we used the function `contest` implemented in the *R* package `r BiocStyle::CRANpkg("lmerTest")`  and used the Satterthwaite [@Kuznetsova2017lmerTest] method to estimate the denominator degrees of freedom. These methods are available in the class `Contrast` (see Figure \@ref(fig:ContrastUML))

@@

# Results and Discussion

@@

The code snippets in this section demonstrate how a DEA workflow can be implemented using the *prolfqua* *R* package. To speed up the computation of these examples, we use a subset of the Ionstar dataset generated by randomly selecting $400$ proteins. First, we remove all proteins with a single peptide and all observations for which *MaxQuant* reports zero intensities which leaves $332$ proteins. Next, peptide abundances are $\log_2$ transformed and robust z-score scaled using the method `robscale`. Then, using the `LFQDataPlotter` class, we show the distribution of the normalized peptide abundances in Figure \@ref(fig:figure1) Panel A. Afterwards, protein intensities are estimated from peptide intensities using Tukey's median polish. Figure \@ref(fig:figure1) Panel B shows the peptide intensities and the estimated protein intensities. Next, we compute the standard deviation of all the proteins in each group and display their distribution using violin plots (Panel C). Finally, we create a boxplot (Panel D) showing the abundance of one protein.

@@

(ref:figure2) Panel A - Histogram showing the distribution of p-values for $332$ proteins for contrasts "e_vs_d" and "d_vs_c". Panel B -  Volcano plot showing $-\log_{10}$ transformed $FDR$ as a function of the difference between groups for $332$ proteins. With black dots, we show effect size and FDR estimates obtained from the linear model, while in green, we plot those obtained using imputation. Finally, panel C shows the difference between groups, as a function of the rank of the abundance of the proteins.

```{r figure2, fig.cap="(ref:figure2)", echo=TRUE, out.width = '90%', fig.width = 8, fig.height = 5}
## specify differences among groups
contrasts <- c(
  "e_vs_d" =   "dilution.e - dilution.d",
  "d_vs_c" =   "dilution.d - dilution.c"
)
## fit model
lmmodel <- paste(lfqpro$response()," ~ dilution.")
modelFunction <- prolfqua::strategy_lm( lmmodel, model_name = "lm")
models <- prolfqua::build_model(lfqpro, modelFunction)

@@

*R* linear model and linear mixed effect models allow modeling parallel designs, repeated measurements, factorial designs, and many more. Models in *prolfqua* are specified using *R*'s linear model and mixed model formula interface. Therefore, knowledge of the *R* regression model infrastructure [@faraway2016extending; @MASS2002] is advantageous when using our package. Furthermore, this glass box approach should make it easy to reimplement an analysis performed with *prolfqua* using base *R* or other programming languages by reading the analysis script.
However, in the package documentation, we showcase how a user, without this knowledge, can analyze experiments with a parallel group design and a factorial design.

@@

To ease the usage barriers of the *R* package to users not familiar with statistics and *R* programming, we developed an application based on the *prolfqua* package into our data management platform [B-Fabric](https://fgcz-bfabric.uzh.ch/) [@bfabric;@bfabricjib2022]. The B-Fabric system runs a computing infrastructure controlled by a local resource management system that supports cloud-bursting [@VMMAD]. This integration enables users to select the input data and basic settings in a graphical user interface (GUI).

In this GUI application, we determine the model formula, and which comparisons to compute from the sample annotation provided in tabular form, similarly to *SAINTexpress* or *MSstats*. However, this application can analyze only parallel group designs with and without repeated measurements or factorial designs without interactions. The user receives a report including input files, the *R* markdown file, and *R* scripts necessary to replicate the analysis using their in-house *R* installation. In this way, *prolfqua* and [B-Fabric](https://fgcz-bfabric.uzh.ch/) help scientists to meet requirements from funding agencies, journals, and academic institutions while publishing their data according to the FAIR  [@FAIR] data principles. We are working on creating a shiny stand-alone application with the described functionality and making it available soon.

@@

The Benchmark functionality of *prolfqua* includes receiver operator curves (ROC) and computes partial areas under those curves ($pAUC$) and other scores, e.g., the false discovery proportion $FDP$. We use those scores, i.e. the $pAUC$ at $10\%$ $FDR$ and the $FDP$, to examine how well the methods implemented in *prolfqua* model quantitative mass spectrometric high throughput data and compare them with the results produced by *MSstats*, *proDA* and *msqrob2*. Table \@ref(tab:describeBenchmarkedModels) summarizes all the modelling methods we evaluated.

@@

Using a benchmark dataset whose ground truth is known (see Methods), we assessed the performance of different modeling approaches implemented in *prolfqua*, *MSstats*, *proDA* and *msqrob2*. Table \@ref(tab:describeBenchmarkedModels) summarizes which methods we have evaluated, which *MaxQuant* files we used as input, and if the models are fitted to peptide or protein intensities. We make the *R*-markdown files to replicate the benchmarking available at `r BiocStyle::Githubpkg("wolski/prolfquabenchmark")`.
